Human Factors Influencing Visual Statistical Inference

### 2013-06-07 sociology_paper.tex
figures for RCP averages
experimental setup updated

\begin{table*}[hbtp] 
\caption{Visual test statistics and question asked to the observers to answer while evaluating a lineup } 
\centering 
\begin{tabular}{m{.5cm}m{3cm}m{3cm}m{6.5cm}} 
\hline\hline 
Serial & Experiment & Test Statistic & Lineup question \\ [0.5ex] % inserts table %heading 
\hline 
1  & Box plot & \begin{minipage}[t]{3cm} \begin{center}	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{center} \end{minipage} & Which set of box plots shows biggest vertical difference 
between group A and B? \\
2 &  Scatter plot & \begin{minipage}[t]{3cm}  \begin{center} \scalebox{0.45}{\includegraphics{stat_beta_k.pdf}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? \\

  3 & Contaminated plot &\begin{minipage}[t]{3cm} \begin{center} \scalebox{0.45}{\includegraphics{stat_contaminated.pdf}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? \\  \\ 
  4 & Polar vs Cartesian & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_polar.pdf}} \end{center} \end{minipage} &Which plot is different? \\ 
  5 & Hist vs density & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_density.pdf}} \end{center} \end{minipage} &In which plot is the blue group furthest to the right?\\ 
  6 & Violine vs boxplot & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_violin.pdf}} \end{center} \end{minipage}&  In which plot does the blue group look the most different from  the red group?\\ 
  7 & Group separation & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.4}{\includegraphics{stat_separation.pdf}} \end{center} \end{minipage} &Which of these plots has the most separation between the coloured groups?\\ 
  8 & Sine Illusion & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.28}{\includegraphics{stat_sine.pdf}} \end{center} \end{minipage} & In what picture is the size of the curve most consistent?\\ 
  9 & Gene expression &\begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.45}{\includegraphics{stat_gene.pdf}} \end{center} \end{minipage} & In which of these plots is the green line the steepest, and the spread of the green points relatively small?\\ 
  10 & Test normality & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_normality.pdf}} \end{center} \end{minipage} &Which of these plots is most different from the others?\\

@@

\section{Simulation Experiment and Methods}\label{sec:exp_design}
The performance of the observer while evaluating a lineup would depend on the factors described in Section \ref{sec:factor_performance}. In this section we present the experimental design to control some  of them so that learning effect of observer and location effect of a plot in the lineup can be estimated.

\subsection{Experiment Setup}  Multiple simulation experiments are designed to investigate the learning trend of the observer and the location effect of the actual data in the lineup. These are explained in tis section.

\paragraph{Design for Learning Trend:} Learning trend of an observer can be observed in terms of performance over successive feedback when multiple lineups are shown for evaluation. Each subject is shown a total of 10 lineups. The lineups are not necessarily with the same difficulty levels. But the order of lineups is randomized. The responses of the lineups are recorded by attempt 1 through 10. Attempt 1 means the response is for the first lineup and attempt 10 means the response is for the 10th lineup. The goal is to estimate whether performance of the observer improves from attempt 1 to attempt 10.

\paragraph{Design for Location effect:} The experiment is set up based on the gene expression data with two groups. The groups are defined based on the presence or absence of iron sufficiency. The null hypothesis is that there is no difference between this two groups. Null plots are obtained by randomizing the group structure of the data. The actual plot is then randomly placed in five different locations. For each location, five different sets of null plots are used to produce five different lineups. This produces 25 different lineups. One such lineup is shown in Figure \ref{fig:lineup_geno}. Similar procedure is applied to obtain 25 lineups to test the interaction between Genotype and Empty Vector (EV).

Each observer is shown three lineups. First lineups shown is a randomly selected Interaction lineup, then one from randomly selected Genotype lineup. Finally a test lineup is shown to screen out unusual evaluation. The subject is not aware which one is the test plot but it is informed in the starting of the task about having a test plot. The test plot is very easy and everyone should detect the data plot. The MTurk worker is paid based on whether he or she could correctly detect the data plot. But the data on test plot are excluded from the analysis.

@@

>>>>

\begin{table*}[hbtp] 
\centering 
\caption{Visual test statistics used in 10 different simulation experiments. The observers are asked different questions to answer while evaluating a lineup} 
\begin{tabular}{m{.5cm}m{3cm}m{3cm}m{6.5cm}} 
\hline\hline 
Serial & Experiment & Test Statistic & Lineup question \\ [0.5ex] % inserts table %heading 
\hline 
1  & Box plot & \begin{minipage}[t]{3cm} \begin{center}	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{center} \end{minipage} & Which set of box plots shows biggest vertical difference 
between group A and B? \\
2 &  Scatter plot & \begin{minipage}[t]{3cm}  \begin{center} \scalebox{0.45}{\includegraphics{stat_beta_k.pdf}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? \\

  3 & Contaminated plot &\begin{minipage}[t]{3cm} \begin{center} \scalebox{0.45}{\includegraphics{stat_contaminated.pdf}} \end{center} \end{minipage} & Of the scatter plots below which one shows data that has steepest slope? \\  \\ 
  4 & Polar vs Cartesian & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.32}{\includegraphics{stat_polar.pdf}} \end{center} \end{minipage} &Which plot is different? \\ 
  5 & Hist vs density & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.38}{\includegraphics{stat_density.pdf}} \end{center} \end{minipage} &In which plot is the blue group furthest to the right?\\ 
  6 & Violine vs boxplot & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_violin.pdf}} \end{center} \end{minipage}&  In which plot does the blue group look the most different from  the red group?\\ 
  7 & Group separation & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.4}{\includegraphics{stat_separation.pdf}} \end{center} \end{minipage} &Which of these plots has the most separation between the coloured groups?\\ 
  8 & Sine Illusion & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.28}{\includegraphics{stat_sine.pdf}} \end{center} \end{minipage} & In what picture is the size of the curve most consistent?\\ 
  9 & Gene expression &\begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.45}{\includegraphics{stat_gene.pdf}} \end{center} \end{minipage} & In which of these plots is the green line the steepest, and the spread of the green points relatively small?\\ 
  10 & Test normality & \begin{minipage}[t]{3cm} \begin{center}  \scalebox{0.35}{\includegraphics{stat_normality.pdf}} \end{center} \end{minipage} &Which of these plots is most different from the others?\\

@@

\section{Simulation Experiment and Methods}\label{sec:exp_design}
The performance of the observer while evaluating a lineup would depend on the factors described in Section \ref{sec:factor_performance}. This section presents the experimental setup, data collection methods and data analysis plans with models to asses the influence of those demographic factors on visual statistical inference.

\subsection{Experiment Setup}  In addition to the evaluation response of the lineup, all the 10 experiments shown in Table \ref{tbl:visual_stat} were designed to collect the following demographic information of the subjects participating the experiments:

\begin{enumerate} \itemsep 0in
\item Age level
\item Gender
\item Academic education level
\item Geographical location
\end{enumerate}

Instead of collecting exact age, 9 levels of age were collected. They are 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, 51-55, 56-60, above 60. There were five levels of academic information. They are (1) High school or less , (2) Some under grad course, (3) Under graduate degree, (4) Some graduate courses,  (5) Graduate degree.  Geographical locations were collected using the true public ip addresses of the participants' computer. This provides latitude and longitude of the ip address as well as the city and country information.

%Multiple simulation experiments are designed to investigate the learning trend of the observer and the location effect of the actual data in the lineup. These are explained in tis section.

\paragraph{Learning Trend:} Learning trend of an observer can be observed in terms of performance over successive feedback when multiple lineups are shown for evaluation. Experiments 5, 6 and 7 were used for this. Each subject was shown a total of 10 lineups randomly selected from a pool of lineups. The lineups are not necessarily with the same difficulty levels. But the order of lineups is randomized. The responses of the lineups are recorded by attempt 1 through 10. Attempt 1 means the response is for the first lineup the observer evaluates and attempt 10 refers to the response for the 10th lineup. The goal is to estimate whether performance of the observer improves from attempt 1 to attempt 10.

\paragraph{Design for Location effect:} The experiment was set up based on the gene expression data with two groups. The groups are defined based on the presence or absence of iron sufficiency. The null hypothesis is that there is no difference between this two groups. Null plots were obtained by randomizing the group structure of the data. The actual plot was randomly placed in five different locations. For each location, five different sets of null plots were used to produce five different lineups. This produced 25 different lineups. Similar procedure was applied to obtain 25 lineups to test the interaction between Genotype and Empty Vector (EV).

%One such lineup is shown in Figure \ref{fig:lineup_geno}.

Each observer was shown three lineups. First lineups shown is a randomly selected Interaction lineup, one from randomly selected Genotype lineup. Finally a test lineup was shown to screen out unusual evaluation. The subjects were not aware which one is the test plot but it was informed in the starting of the task about having a test plot. The test plot was very easy and everyone should detect the data plot. The MTurk worker was paid based on whether he or she could correctly detect the data plot. But the data on test plot are excluded from the analysis.

@@

\subsection{Demographic Factors}
